title: "Bio 723: Assignment 05 -- Matrices and multiple regression"
author: "Chelsea Shoben CRS70"
output:
  html_document:
    highlight: default
    fig_caption: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, eval = TRUE, 
                      warning = TRUE, comment=NA, cache = FALSE,
                      fig.width = 6, fig.height = 4)
```

Bio 723: Assignment 05


#Problems
Each of the problems below is worth 4 pts.

### 1. Write a function, mean.vector(X), that takes a matrix of real values (doubles) and calculates the column means.

```{r}

A <- matrix(1:16, nrow=4)

```



```{r}

mean.vector <- function(x) {
  k <- 1/(nrow(x))
  one.vector <- matrix(1:1, nrow=nrow(x))
  m <- k * (t(one.vector) %*% x)
  
  
m

  
  #mean.list <- rep(NA, ncol(x))
  #for (i in 1:ncol(x)) {
   # mean.list[i] <- mean(x[,i])
  #}
#mean.list
}
A
mean.vector(A)

```







### 2. Write a function, deviation.matrix(X), that takes a matrix of real values and returns a matrix the represents each values deviation from it’s respective column mean.



```{r}
deviation.matrix <- function(x) {
  one.vector <- matrix(1:1, nrow=nrow(x))
  M <- one.vector %*% mean.vector(x)
  x - M
  
}
deviation.matrix(A)
```







### 3. Write a function covariance.matrix(X), that takes as input a matrix of real values (variables in columns) and returns a covariance matrix.

```{r}
covariance.matrix <- function(x) {
  n <- nrow(x) - 1 #this is like the n-1 in the denominator
  k <- 1/n #this will be used to multiply the D^T, D dot product
  D <- deviation.matrix(x) #The deviation matrix of the matrix. 

  S <- k * (t(D) %*% D)
S


  
}

covariance.matrix(A)
cov(A) # to check my answer
```






### 4. Write a function correlation.matrix(X), that takes as input a matrix of real values (variables in columns) and returns a correlation matrix.

```{r}
correlation.matrix <- function(x) {
  diag.vector <- diag(covariance.matrix(x)) #make a vector containing only the diagonal items from the matrix. 
  Sii <- 1/sqrt(diag.vector)
  V <- diag(Sii)
  V %*% covariance.matrix(x) %*% V
  }
correlation.matrix(A)
cor(A) #check my answer


```









### 5. Write a function multiple.regression(X, y), that takes as input a matrix of real values (X; the explanatory variables in columns) and a real-valued vector y (the dependent variable), and which returns a list giving: 
### 1) the regression coefficients for the multiple regression of y on X, and 
### 2) the coefficient of determination for the regression.




```{r}
#needed to solve below
B <- matrix(1:4, nrow=2)
C <- c(1:4)

vec.dot <- function(x,y) {
  products <- x * y
  sum(products)
}

vec.length <- function(x){
  sqrt(vec.dot(x,x))

}
B  

```

```{r}
solve(B)
solve(t(B) %*% B)
```




```{r}
multiple.regression <- function(x,y) {
  holding <- solve(t(x) %*% x) #((X^T)X)^1
  b <- (holding %*% t(x)) * y
  
  y.hat <- x %*% b
  length.y.hat <- vec.length(y.hat)
  length.y <- vec.length(y)
  R <- (length.y.hat) / (length.y)
  
list('#1 regression coefficients', b, '#2 coefficient of determination', R)
}

multiple.regression(B, C)
```















