---
title: "Bio 723: Assignment 06 -- More on Regression"
author: "Chelsea Shoben CRS70"
output:
  html_document:
    highlight: default
    fig_caption: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, eval = TRUE, 
                      warning = TRUE, comment=NA, cache = FALSE,
                      fig.width = 6, fig.height = 4)
```



#Bio 723: Assignment 06

Instructions
For each of the problems below (except in cases where you are asked to discuss your interpretation) write R code blocks that will compute appropriate solutions. A good rule of thumb for judging whether your solution is appropriately “computable” is to ask yourself “If I added or changed observations to this data set, would my code still compute the right solution?”

When completed, submit your R Markdown document (the file with the extension .Rmd) via Sakai.

#Make your solutions computable
For each of the problems below (except in cases where you are asked to discuss your interpretaion) write R code blocks that will compute appropriate solutions. A good rule of thumb for judging whether your solution is appropriately “computable” is to ask yourself “If I added or changed observations to this data set, would my code still compute the right solution?”

When completed, submit your R Markdown document (the file with the extension .Rmd) and the knit HTML via Sakai.

#Problems
#Bivariate regression

#1.Using the lion data, fit a regression model relating age to the proportion of black pigmentation. Augment the lion data frame using the broom::augment function as described in the workbook. Then answer the following questions:

```{r}
#libraries

library(tidyverse)
library(ggExtra)
library(broom)

lions <- read_csv("https://github.com/bio304-class/bio304-course-notes/raw/master/datasets/ABD-lion-noses.csv")

```

```{r}
ggplot(lions, aes(x = proportionBlack, y = ageInYears)) + 
  geom_point() +
  geom_smooth(method="lm", color = 'red')
```



```{r}
fit.lion <- lm(ageInYears ~ proportionBlack, data = lions)
fit.lion
```



```{r}
lion.augmented <- augment(fit.lion, lions)
ggplot(lion.augmented, aes(proportionBlack, .resid)) +
  geom_point()+
  geom_hline(yintercept=0, color='firebrick', linetype='dashed')
```

```{r}

head(lion.augmented[order(lion.augmented$proportionBlack, decreasing = FALSE),])
```





##1a. Based on the regression of age on proportion of black pigmentation, if you observed a male lion with no black pigmentation on its nose, how old would you predict it to be? [1 pt]

A little over a year old (+ or - a year)


##1b. What is the predicted increase in age for a 10% increase in black pigmentation? [1 pt]

1.25 years


##1c. Calculate the interquartile range (IQR) for the residuals of the model fit [1 pt]

```{r}
lion.resid <- lion.augmented$.resid
lion.resid
IQR.r <- IQR(lion.resid)
IQR.r
```



## 1d. Use dplyr::mutate to add another column called resid.outlier to lions.augmented. resid.outlier should be TRUE if the absolute value of the corresponding residual is more than 1.5 * IQR [1 pt]


```{r}
typeof(lion.augmented)
class(lion.augmented)
```

```{r}
head(lion.augmented)
```


```{r}
lion.augmented.plus <- mutate(lion.augmented, resid.outlier =  NA)   #Make an empty column.        
```


```{r}
boolean.resid <- ifelse (abs(lion.resid) > (1.5*IQR.r), TRUE,FALSE) #make BOOLEAN to put into the empty column created.

```


```{r}
lion.augmented.plus$resid.outlier <- boolean.resid

head(lion.augmented.plus)

```


##1e. Use the resid.outlier column from the previous question to draw a plot like the one below, highlighting potential outliers in regression model. [1 pt]

```{r}
ggplot(lion.augmented.plus, aes(x = proportionBlack, y = ageInYears, color=resid.outlier)) + 
  geom_point() +
  geom_smooth(method="lm", color = 'red')+
  labs(x = 'Proportion Black Pigmentation', y = 'Age(years)', title = 'Regression of Age on Nose Pigmentation for Male Lions')
```




#Multiple regression
The HistData package is required to have access to the data for this assignment. Install HistData via one of the standard mechanisms.

The GaltonFamilies data set in the HistData package lists observations for heights of parents and their adult offspring for 934 children in 205 families, a famous data set collected by Francis Galton (Galton, F. (1886). Regression Towards Mediocrity in Hereditary Stature. Journal of the Anthropological Institute, 15, 246-263).




#2 Load the HistData package and examine the GaltonFamilies data set.  

```{r}
library(HistData)
summary(HistData::GaltonFamilies)
head(HistData::GaltonFamilies)
```

# 2 cont. Using cowplot create a single figure with three sub-figures:
## A) the distribution of height for all offspring;
## B) overlapping density plots giving the height distributions for male and female offspring separately; 
## C) boxplots of height for male and female children separately [3 pts]

```{r}
library (ggplot2)
library (cowplot)

```

```{r}
gg_hist_childHeight <- ggplot(GaltonFamilies) +
  geom_histogram(aes(x=childHeight), bins = 15) +labs(title = 'Distribution of Child Heights', x = 'Child Height') + theme(plot.title = element_text(size = 10))

gg_density_childHeight <- ggplot(GaltonFamilies) +
  geom_density(aes(x=childHeight, fill = gender), alpha=0.25) + 
  labs(title='Child Height Distribution \n by Gender', x = 'Child Height', fill='Gender') + 
  theme(plot.title = element_text(size=10))



gg_box_childHeight <- ggplot(GaltonFamilies) +
  geom_boxplot(aes(x = gender, y = childHeight, color = gender)) +
  labs (title='Child Height Boxplot \n by Gender', x = 'Gender', y = 'Child Height', fill = 'Gender') +
  theme(plot.title = element_text(size=10))


plot_grid(gg_hist_childHeight, gg_density_childHeight, gg_box_childHeight, ncol=2, labels=c('A', 'B', 'C'))
```


#3 Create two 3D scatter plots, one each for male and female offspring, showing the relationship between offspring height and mother and fathers height. [2 pts]

```{r}
library(tidyverse)
library(broom)
library(scatterplot3d)
```

```{r}
femalechild.galtonfamilies <- GaltonFamilies[GaltonFamilies$gender == 'female',]

malechild.galtonfamilies <- GaltonFamilies[GaltonFamilies$gender == 'male',]

scatterplot3d::scatterplot3d(femalechild.galtonfamilies$mother, femalechild.galtonfamilies$father, femalechild.galtonfamilies$childHeight, angle=40, main='Female Child Height Relative to\n Mother and Father Height', xlab = 'Mother Height', ylab = 'Father Height', zlab = 'Female Child Height', box = FALSE )
```

```{r}
scatterplot3d::scatterplot3d(malechild.galtonfamilies$mother, malechild.galtonfamilies$father, malechild.galtonfamilies$childHeight, angle=40, main='Male Child Height Relative to\n Mother and Father Height', xlab = 'Mother Height', ylab = 'Father Height', zlab = 'Male Child Height', box = FALSE )
```


#4 For male offspring, fit a multiple regression of offspring height on father and mother's height. Write out the predicted model with the corresponding coefficients in the form Oˆmale=a+b1F+b2M where O, F and M are offspring, father's, and mother's height. What fraction of the variation in offspring height does the model capture? [2 pts]

```{r}
male.regression <- lm(childHeight ~ father + mother, data = malechild.galtonfamilies)
male.regression
summary(male.regression)
```

Based on the R-squared value, it captures 23.47%


#5 For female offspring, fit a multiple regression of offspring height on father and mother's height. Write out the predicted model with the corresponding coefficients in the form Oˆfemale=a+b1F+b2M where O, F and M are offspring, father's, and mother's height. What fraction of the variation in offspring height does the model capture? [2 pts]

```{r}
female.regression <- lm(childHeight ~ father + mother, data = femalechild.galtonfamilies)
female.regression
summary(female.regression)
```
According to the R-squared value it captures 26.49% of the variance. 



# 6. What is the predicted height of a male child, if the mother was 5 ft tall and the father was 6 ft tall? What is the predicted height of a female child from the same parents? [2 pt]

```{r}
father.inches <- 6*12
mother.inches <- 5*12

predicted.male <- 19.3128 + 0.4176*father.inches + 0.3188*mother.inches

predicted.female <- 18.8336 + 0.3725*father.inches + 0.3035*mother.inches

predicted.male
predicted.female
```


LOESS regression


# 7. Write an R function that genereates a plot that simultaneously illustrates trends in both approval and disapproval ratings for Barack Obama, showing both the raw data and corresponding LOESS fits. Use colors and/or shapes to distinguish the two trends. Make sure both your x- and y-axes are scaled to show the full range of the data. Label your axes and create a title in the plot. Aim for a 'publication quality' figure [3 pts]

```{r}
polls <- read_delim('https://github.com/Bio723-class/example-datasets/raw/master/obama-polls-2008-2011.txt',delim="\t", trim_ws=TRUE)
```

```{r}
#for some reason it would not knit the code below. It was running fine on my end but chose to comment it out for submission purposes. 

#polls.app.and.dis <- 
  #polls %>%
  #ggplot(aes(x = date)) + 
  #geom_point(aes(y = Approve), color='firebrick', alpha=0.5, pch=1) +
  #geom_point(aes(y = Disapprove), color = 'blue', alpha=0.5, pch=1) + 
  #geom_smooth(aes(y= Approve), color='red', method="loess", se=FALSE) +
  #geom_smooth(aes(y = Disapprove), color='royalblue', method="loess", se=FALSE) +
  #labs(x = 'Date', y = 'Opinion', title = 'Barack Obama Opinion Ratings, 2008-2012') +
  #theme_minimal() 

#polls.app.and.dis
  
```


Logistic regression

Bumpus (1898) described a sample of house sparrows which he collected after a very severe storm. The sample included 136 birds, sixty four of which perished during the storm. Also included in his description were a variety of morphological measurements on the birds and information about their sex and age (for male birds). This data set has become a benchmark in the evolutionary biology literature for demonstrating methods for analyzing natural selection. The bumpus data set is available from the class website as a tab-delimited file bumpus-data.txt.

Reference: Bumpus, H. C. 1898. The elimination of the unfit as illustrated by the introduced sparrow, Passer domesticus. (A fourth contribution to the study of variation.) Biol. Lectures: Woods Hole Marine Biological Laboratory, 209-225.

```{r}
birds <- read_delim('https://github.com/bio304-class/bio304-course-notes/raw/master/datasets/bumpus-data.txt', delim ='\t', trim_ws =TRUE)
```

```{r}
head(birds)
```


# 8. Measures like weight and skeletal dimensions are often proxies for body size. Fit logistic regression models for: a) survival as a function of weight; and b) survival as a function of skull width and report the coefficients of the models. [3] pts]


```{r}
birds.boolean <- as.numeric(birds$survived)
birds.boolean #let's look at it. 
birds.numeric <- birds #make a copy to mutate
birds.numeric$survived <- birds.boolean
head(birds.numeric)

```



```{r}
fit.survival.weight <- glm(survived ~ weight.g, family = binomial, birds.numeric)

fit.survival.skull <- glm(survived ~ width.skull.in, family=binomial, birds.numeric)

tidy(fit.survival.weight)
tidy(fit.survival.skull)
```
The 'estimates' above are the coefficients. 


# 9. Illustrate the two regression models from the previous question using scatter plots overlain with logistic regression curves. Combine the two plots as subplots A) and B) in a single figure, using cowplot. [3 pt]

```{r}
#make the scatter plots
weight.plot <- ggplot(data = birds.numeric, aes(x = weight.g, y = survived)) +
  geom_jitter(width = 0.1, height = 0.05, color = 'pink') +
  labs(title = 'Bird Survival as a Function of Weight') +
  theme(plot.title = element_text(size=10))

skull.plot <- ggplot(data = birds.numeric, aes(x = width.skull.in, y = survived)) +
  geom_jitter(width = 0.1, height = 0.05, color = 'green') +
  labs(title = 'Bird Survival as a Function of\n Skull Width') +
  theme(plot.title = element_text(size=10))


```



```{r}
weight.g <- seq(21, 32, 0.1)

predicted.s.weight <- predict(fit.survival.weight, newdata = data.frame(weight.g = weight.g), type = 'response')


s.logistic.weight.plot <- weight.plot + geom_line(data = data.frame(weight.g = weight.g, survived = predicted.s.weight), color = 'darkgreen', size = 1)

width.skull.in <- seq(0.55, 0.66, 0.001)

predicted.s.skull <- predict(fit.survival.skull, newdata = data.frame(width.skull.in = width.skull.in), type = 'response')


s.logistic.skull.plot <- skull.plot + geom_line(data = data.frame(width.skull.in = width.skull.in, survived = predicted.s.weight), color = 'darkblue', size = 1)



plot_grid(s.logistic.weight.plot, s.logistic.skull.plot)
```



# 10. Use dplyr::group_by and dplyr::do() to fit logistic regressions of survival as a function of weight for male and female birds separately, using broom::tidy to return a table of coeffficients (and other parameters) from the model fits [3 pts]

```{r}


grouped.birds <- 
  birds.numeric %>% 
  group_by(sex) %>%
  do(fits = glm(survived ~ weight.g, family = binomial, data = .))

tidy(grouped.birds$fits[[2]])
class(grouped.birds)
typeof(grouped.birds)
```



# 11. Using faceting to produce a figure illustrating the logistic regression of survival as a function of weight for male and female birds separately [2 pt]


```{r}
faceted.plot.sex <- ggplot(birds.numeric, aes(x = weight.g, y=survived, color = sex)) +
  geom_jitter(width = 0, height = 0.06) +
  geom_smooth(method='glm', method.args = list(family='binomial')) +
  labs(x = 'Weight in Grams', y = 'P(Survived)') +
  facet_wrap(~sex)
```

```{r}

weight.g <- seq(21, 32, 0.1)


predicted.grouped.birds <- predict(grouped.birds, newdata = data.frame(weight.g = weight.g), type = 'response')


final.plot <- faceted.plot.sex + 
  geom_line(data = data.frame(weight.g = weight.g, survived = predicted.grouped.birds))
```

I was unable to knit due to the error above. I was not able to figure out which class predit() needed it to be to run the code. 












