---
title: "Shoben_C_Assignment04"
output:
  html_document:
    df_print: paged
---

---
title: "Bio 723: Assignment 04 -- Variables as vectors"
author: "Chelsea Shoben CRS70"
output:
  html_document:
    highlight: default
    fig_caption: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, prompt = FALSE, eval = TRUE, 
                      warning = TRUE, comment=NA, cache = FALSE,
                      fig.width = 6, fig.height = 4)
```




# Bio 723: Assignment 04

Instructions
For each of the problems below (except in cases where you are asked to discuss your interpretation) write R code blocks that will compute appropriate solutions. A good rule of thumb for judging whether your solution is appropriately “computable” is to ask yourself “If I added or changed observations to this data set, would my code still compute the right solution?”

When completed, submit your R Markdown document (the file with the extension .Rmd) via Sakai.

Make your solutions computable
For each of the problems below (except in cases where you are asked to discuss your interpretaion) write R code blocks that will compute appropriate solutions. A good rule of thumb for judging whether your solution is appropriately “computable” is to ask yourself “If I added or changed observations to this data set, would my code still compute the right solution?”

When completed, submit your R Markdown document (the file with the extension .Rmd) via Sakai.

#Problems
Each of the problems below is worth 2.5 pts.

### 1. Write a function, vec.dot(x,y), that calculates the dot product of two vectors, x and y, of arbitrary number of elements.


```{r}
vec.dot <- function(x,y) {
  products <- x * y
  sum(products)
}

#examples

x <- c(1, 2, 3)
y<- c(4, 5, 6)


vec.dot(x, y)
```









### 2. Write a function, vec.length(x), that calculates the Euclidean length of a vector. Your function should call your vec.dot() function aove.

```{r}
vec.length <- function(x){
  sqrt(vec.dot(x,x))
}
  
vec.length(x)

```









### 3. Write a function vector.cosine(x,y), that calculates the cosine of the angle between two vectors.

```{r}
vector.cosine <- function(x, y) {
  (vec.dot(x,y)) / (vec.length(x) * vec.length(y))
}

vector.cosine(x,y)


```




### 4. Write a function vector.angle(x,y, deg=FALSE) that by defaults calculates the angle, in radians, between two vectors. When the argument  deg=TRUE your function should return the angle in degrees.

```{r}
vector.angle <- function(x,y, deg=FALSE) {
  radian <- acos(vector.cosine(x,y))
  degrees <- (radian*180)/pi
  ifelse ((deg==FALSE), radian, degrees)

}

vector.angle(x,y)
vector.angle(x, y, deg=TRUE)
```





### 5. Write a function, vector.distance(x,y) that calculates the distance between two vectors (i.e. distance between their end points).

```{r}
vector.distance <- function(x,y) {
  sqrt(vec.dot(x,x) + vec.dot(y,y) - (vec.dot((2*x),y)))
}

vector.distance(x,y)
```





### 6. Write a function vector.mean(x) that calculates the mean of x, using vector algebraic operations.

```{r}
vector.mean <- function(x) {
  n <- length(x)
  sum <- sum(x)
  (1/n) * sum
}

vector.mean(x)
```





### 7. Write a function vector.cov(x,y) that calculates the covariance of x and y, using vector algebraic operations. Remember to mean center!


```{r}
#mean centering
mean.center <- function(x) {
  x - vector.mean(x)
}

mean.center(x)
mean.center(y)


```



```{r}
vector.co <- function(x,y) {
  X <- mean.center(x)
  y <- mean.center(y)
  ((vec.dot(x,y)) / (length(x)-1))
}

vector.co(x,y)
```





### 8. Write a function vector.var(x) that calculates the variance of x, using vector algebraic operations. Remember to mean center!

```{r}
vector.var <- function(x) {
  x <- mean.center(x)
  (vec.dot(x,x)) / (length(x) -1)
}

vector.var(x)
```





### 9. Write a function vector.cor(x,y) that calculates the correlation of x and y, using vector algebraic operations.Remember to mean center!

```{r}
vector.cor <- function(x,y) {
  (vec.dot(mean.center(x),mean.center(y))) /
     (vec.length(mean.center(x)) * vec.length(mean.center(y)))
}
vector.cor(x,y)
```





### 10. Write a function vector.regression(x,y) that calculates the slope of the regression of y on x using vector algebraic operations. Remember to mean center!

```{r}
vector.regression <- function(x,y) {
  (vec.dot(mean.center(x), mean.center(y))) / (vec.dot(mean.center(x), mean.center(x)))
  
}

vector.regression(x,y)
```









