---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

Data codebook:  https://tinyurl.com/2p8snxdf
Project Goals:  https://tinyurl.com/pjyjpzbw

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library (tidyr)
library (tidyverse)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data


Data contained in the BRFSS (Behavioral Risk Factor Surveillance System) is through an observational study that collects information about non-institutionalized adults (18 years +) across the US by state. This information is collected through telephone interviews of an adult randomly selected from a household. While I was not able to fi nd documentation on whether or not the household called was randomly (was it clustering to help with generaliziability?) selected I was able to fi nd that the definition of a household and that college housing was included. Cellular phones also stand as a single-person household if called for an interview.
While there may be an attempt for random sampling, I think this study has too many factors that will impact the attempt at being random. First, the non-response bias of a phone interview survey must be substatial. From my personal experiences, I predict more people are going to hang-up rather than answer the questions. Additionally, there is nothing random about the adult of a household who answers the phone. It is likely the household is bias where one adult is more willing or more available to answer.That would ultimately skew these results. Additionally, some adult groups are more likely to physically be home during the call(those who work from home, work unpredictable hours, or are retired) and this again reduces the random nature of the samples. This study does not capture the data from individuals who may have group housing or displaced housing, and possibly with a single shared house phone and no cell phones. These are all biases in the person who answers the phone. Additionally,there are biases in the type of people willing to answer. Some personalities are going to be more willing to coorporate. Any individual whose first language is not English is more likely to refuse or misconceive the question being asked. There are likely sex, gender, profession, life history, and a number of other factors that will reduce the randomness of these results.
The results of this study can only be used for generalized associations between the samples. And with that, not a robust generalization of the population of interest, but rather simply the population who answers the phone and completes the survey.No causality can be inferred because variables are not controlled and individuals were not assigned to random groups and then studied.




* * *

## Part 2: Research questions

**Research quesion 1:**

Is there a relationship between sex, race, and number of work hours? Spefi cially, a relationship betweenrace, sex, and individuals that work overtime (more than 40 hours per week)?

Often individuals from low-income communities work multiple part-time jobs. Their combined hours can amount to well above the average full-time job. Minority races disproportionately make up a large fraction of the lower-class. I'm interested in seeing ifdata from these phone interviews supports this narrative?


**Research quesion 2:**

Is there an association between race, health status, and health care coverage?
Lacking health care coverage while identifying as having a low health status is a bad situation. Is there a bias across races that identify in this way?

**Research quesion 3:**

Is there a relationship between race and having high blood pressure (hypertension) during pregnancy forfemales in this study?

Pre-eclampsia is more likely to occur in Black pregnant women than in White pregnant women. This can lead to the death of the child, mother, or both. Does this data show evidence of that?
https://www.heart.org/en/news/2019/02/20/why-are-black-women-at-such-high-risk-of-dying-from-pregnancy-complications https://www.heart.org/en/news/2019/02/20/why-are-black-women-atsuch-
high-risk-of-dying-from-pregnancy-complications
https://nortonhealthcare.com/news/pregnant-african-american-women-pre-eclampsia/# 

* * *

## Part 3: Exploratory data analysis


**Research question 1:**

Is there a relationship between sex, race, and number of work hours?

Variables: sex: sex of respondent; levels: Male Female scntwrk1: How Many Hours Per Week Do You Work (1-96 are hour, 97 -Don't know/not sure, 98 - zero, 99 - refused)) X_mrace1 - race of individual (excluding ethnicity data)
I am going to limit this analysis to respondents that work more than 40 hours per week, simply because 1) the data is too large otherwise and 2) I think this will still be interesting to consider.

#### Q1 Selecting and cleaning data

```{r}
# 1) Make new data frame with only variables of interest
q1.work = brfss2013 %>%
select (sex, scntwrk1, X_mrace1)

# 2) Remove rows with NAs.
q1.work = q1.work %>%
drop_na()

# 3) Drop levels for easier manipulation later.
q1.work <- droplevels(q1.work)
# 4) Rename the work column to make it easier to remember.
q1.work = q1.work %>%
rename(hoursworked = scntwrk1)

#5) Remove value of 97, 98, 99 from the work columns which refer to "dont know", "zero", or "refused" respectively.

q1.work = q1.work[!(q1.work$hoursworked == 97),]
q1.work = q1.work[!(q1.work$hoursworked == 98),]
q1.work = q1.work[!(q1.work$hoursworked == 99),]

# 6) Filter to only include those who worked over 40 hours per week.
q1.filter <- filter(q1.work, hoursworked > 40)

# 7) Overall summary for the reader
summary(q1.filter)
```

#### Q1 Graph

```{r}
ggplot(
  q1.filter, 
  aes(x = X_mrace1, y = hoursworked, fill = sex)) +
geom_boxplot(position = "dodge2") +
scale_x_discrete(guide =
guide_axis(
n.dodge=5)) +
labs( 
  title= "Respondant's Race, Sex, and Hours Worked per Week",
  subtitle = "Race & Sex vs. Hours Worked (starting at 41 hours per week)") +
xlab("Race") +
ylab ("Hours Worked per Week")
```

#### Q1 Conclusion

Considering those that work more than 40 hours per week as "over time". It appears that mean number of hours worked perweek (for those who work overtime), even when stratifi ed by sex, are similar between White, Black, Asian, NativeHawaiian/Pacifi c Islander, and other race. A diff erence in sexes is seen for American Indian or Alaskan Native, with femaleaverage work hours being higher. Female American Indian or Alaskan Native seem to have the highest skew of overtime hours,with all responses being captured in the whiskers of the plot (no outliers). In other words, the third quartile range for NativeAmerican or Alaskan Native females spans well into the range of outliers for all other groups, indicating that this high number ofwork hours is more common for this group. Average work hours for multiracial males that worked overtime was higher thantheir female counterparts. I think that it would be useful if the survey asked if individuals worked more than one job and I expect we would have seen atrend across both race and sex. I also wonder how many people this missed, especially those that work overtime as I imaginethey would have been working when these interviews were conducted. I also wonder about how this data is skewed bypriviledged races that are more likely to have a salary job or a job with fl exibility where they could step away for a call during theday. I personally have worked multiple jobs in food and did not have the ability to just walk away when needed. These types ofjobs (essential workers) are more likely to be fi lled by more marginalized races and thus more likely to have been missed duringthis data collection.
Other limitations to this interpretation include what I have already expressed at the beginning of this doc (sample bias, responsebias, use of contact/data collection bias, etc.)
I do not believe these results can be generalized to the American people as a whole, and further data would need to be collected(randomized sampling) to speculate further.


**Research question 2:**

Is there an association between race, health status, and health care coverage?
X_mrace1: race of individual GENHLTH: 1 excellent, 2 very good, 3 good, 4 fair, 5 poor, 7 dont know, 9 refused. HLTHPLN1: healthcare coverage. 1- Yes, 2- No, 7- dont know 9- refused
Is there a bias by race for having poor health and no health care coverage?

#### Q2 Selecting and cleaning data

```{r}

# 1) Select for the three variables of interest, remove NAs, and drop levels
q2.data =
  brfss2013 %>%
select(X_mrace1, genhlth, hlthpln1)
q2.data = q2.data %>%
drop_na()

# 2) Filter for the respondents that meet the criteria of interest (low health and no health insurance).
keep <- c("Poor")
q2.filter <- filter(q2.data, genhlth%in%keep)
q2.filter <- filter(q2.filter, hlthpln1 == "No")

# 3) Need to normalize if I am going to look across races. My goal is to divide the total number of individuals from each race in q2.filter by the total number of individuals for each race that answered both the question on general health and health insurance (q2.data). Then multiply that value by 100 to get the percentage of respondents fitting my criteria from each race category. I will round these percentages for easy graphing later.
totalRaceQ2 <- count(q2.data, X_mrace1)
races <- data.frame(totalRaceQ2$X_mrace1)
filterRaceQ2 <-count(q2.filter, X_mrace1)
tot.race <- data.frame(
  Race = races$totalRaceQ2.X_mrace1,
  Total = totalRaceQ2$n,
  Filter = filterRaceQ2$n)
tot.race$Normalized <- tot.race$Filter/tot.race$Total
tot.race <- tot.race %>% mutate(Percentage =Normalized*100)
tot.race$PercentageRound = round(tot.race$Percentage, digit =2)

# 4) Summary of final data frame
summary(tot.race)

```



**Research quesion 3:**

```{r}

```

